---
title: "Analysing Plankton Diet Metabarcoding Data"
author: "Andreas Novotny"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Analysing Plankton Diet Metabarcoding Data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r message=FALSE}
# Loading packages required for this vingette
library(dada2)
library(phyloseq)
library(DESeq2)
library(tidyverse)
library(metaphylo)
```

# 1. Opening data and construct a phyloseq object

This is the RAW output data files from the dada2 pipeline analysis at UPPMAX, together with the metadata. Nothing has been modified.

```{r message=FALSE}
#Loading data from this package, to be used in the vingette
data("metadata_16S")
data("seqtab_16S")
data("tax_16S")
```



Use help documentation to see more information about data and functions:

```{r}
?metadata_16S
?filterTaxonomy
```


### 1.1 Modify the taxonomy output from dada2

The *filterTaxonomy* function deletes all the taxonomic information with high uncertainty,
below a certain threshold for the bootstrap value, and replaces it with Na.

To avoid Na`s in the data set, that carries no information, the *completeTaxonomy* function fills the empty gaps with information from the higher taxonomic levels.

```{r}
tax_16S <- filterTaxonomy(tax_16S, minboot = 75)
tax_16S <- completeTaxonomy(tax_16S, string = "_x")
```




### 1.2 Build a phyloseq object

The function *dada2ToPhyloseq* takes the specific output from the dada2 pipeline as input to create a phyloseq object.
Tax glom is a phyloseq function that collapses sequence variants with the same taxonomic affiliation, in this case at species level (highest taxonomic resolution). 

*This operation might take up to 5 minutes*.

```{r eval=FALSE}
ps_16S <- dada2ToPhyloseq(seqtab_16S, tax_16S, metadata_16S) %>% 
  tax_glom(taxrank = "Species")
```

### 1.3 Create a CSV table from the phyloseq object, to open in excel.

At any time, any created and modified phyloseq object can be saved as a csv file.
```{r eval=FALSE}
phyloseqToCSV(ps_16S, header="ID_friendly", filename="~/Desktop/RSV_16S.csv")
```

```{r}
data(ps_16S)
```


# 2. Working with the phyloseq object

*This is short summary of the essential functions of the phyloseq package only*

The phyloseq object created in the first section can be loaded directly in this package:

```{r}
data(ps_16S)
data(ps_18S)
```

### 2.1 Browse through the phyloseq object:

The phyloseq object consists of three data tables: the OTU_table (with abundance data), the sample_data, (containing the metadata), and the tax_table (containing the taxonomic information).

```{r}
ps_16S
```

Each data table can be accessed with the @ symbol:
```{r eval=FALSE}
ps_16S@sam_data
```


These 4 functions gives access to specific information the phyloseq object
```{r eval=FALSE}
sample_names(ps_16S) # Formal names of samples

sample_variables(ps_16S) # Metadata variables

rank_names(ps_16S) # Taxonomic levels

get_variable(ps_16S, "SORTED_type") # Access any metadata variable
```


### 2.2 Filter the phyloseq object:

**Easy filtering is the most important feature of phyloseq:**

*subset_samples* allows for sub-setting by column in the metadata:

```{r eval=FALSE}
# To get samples from August:
ss <- subset_samples(ps_16S, MONTH=="aug")
get_variable(ss, "ID_friendly")

#To get only rotifers:
ss <- subset_samples(ps_16S, SORTED_type=="Rotifer")
get_variable(ss, "ID_friendly")
```


*subset_taxa* allows for sub-setting by the taxonomic level:

```{r eval=FALSE}
# To get only the chloroplast taxonomy
ss <- subset_taxa(ps_16S, Domain=="Chloroplast")
ss # Only 100 taxa left
```

*filter_taxa* allows for filtering by the abundance data. This is done by using a logical function(x).
The function goes through the OTU table, taxa by taxa. The function can be simple or complicated:

```{r eval=FALSE}
# sum(x) has to be bigger than 0, ie taxa with no counts are discarded.
ss <- filter_taxa(ps_16S, function(x) sum(x)>0, TRUE)
ss
```

*transform_sample_counts* allows for data transformation, by using a logical function(x).

```{r eval=FALSE}
# To make log transformation:
tps <- transform_sample_counts(ps_16S, function(x) log(x))

# Percentage of full library:
tps <- transform_sample_counts(ps_16S, function(x) x/sum(x)*100)
```


Finally these functions can be combined to access the exact data-set wanted.
In this example the cyanobacteria from Synchaeta in August are visualized as percentage:

```{r}
ss <- ps_16S %>% # We use our phyloseq object as an input
  subset_samples(SORTED_genus=="Synchaeta") %>%  # First we select only rotifers
  subset_samples(MONTH=="aug") %>%  #Then we select only samples from august
  subset_taxa(Phylum=="Cyanobacteria") %>%  # We are only interested in the Cyanobacteria
  transform_sample_counts(function(x) x/sum(x)*100) %>%   # Counts are transformed into percentage
  plot_bar(fill="Genus") # And finally ploted
ss
```



# 3. Analysing the phyloseq object with DEseq2

**OBS** Data going to any DESeq2 function should never be transformed or normalized!

**VST** is Variance Stabilizing Transformation and is conducted by the DESeq2 functions. This is the proper transformation to be used for metabarcoding data (assumes negative binomial distribution).


### 3.1 Get table of significant fold change expressions

This code will compares the diet of Synchaeta and Keratella.
*deseqTable* sends the phyloseq object to DESeq2 and compares two groups.
The "~" statement is the key of the experimental design.
```{r eval=FALSE}
?deseqTable
```

```{r eval=FALSE}
dds_table <- ps_16S %>% 
  subset_samples(MONTH == "aug") %>%
  subset_samples(SORTED_type == "Rotifer") %>% 
  deseqTable(~ SORTED_genus)
dds_table
```

### 3.2 Select relevant prey organisms and make boxplot

*relevantSpecies* Is a way to select for "frequency of occurrence". Here the "cutoff" has an essential function.

*abundancePlot* is a way to plot the abundance of different taxa among replicates of samples. (Uses Box plot of ggplot)

```{r eval=FALSE}
?relevantSpecies
?abundancePlot
```


```{r message=FALSE}

plot <- ps_16S %>% 
  subset_samples(SORTED_genus == "Synchaeta") %>% # Select synchaeta
  relevantSpecies(method = "vst", cutoff = 0.75, design = ~ ID_friendly) %>% # Taxa that are present in less than 70% of the samples are filtered away.
  abundancePlot(x = "Species", fill = "Genus")
plot

```


### 3.3 Create heat map of significant and relevant fold change expressions

*getSignificant* is a way to filter a phyloseq object so that only significantly different taxa remains. It uses DESeq2 to get significance values.
Synchaeta from august and march is compared.


First we make a list of relevant taxa, that are present in many of the samples:
```{r message=FALSE}

relevant_otu <- ps_16S %>% 
  subset_samples(SORTED_genus=="Synchaeta") %>% # Synchaeta is selected
  subset_samples(MONTH!="may") %>% # Synchaeta from may is removed.
  relevantSpecies(method = "vst", cutoff=0.3, design = ~MONTH) %>% # Taxa has to be present in more than 30% of the samples.
  taxa_names()
```

Then we use the list to select only significant taxa:
Here, again the "~" is the important experimental design. We are comparing the effect of MONTH.

```{r message=FALSE}
plot <- ps_16S %>%
  subset_samples(SORTED_genus=="Synchaeta") %>%
  subset_samples(MONTH!="may") %>%
  prune_taxa(relevant_otu, .) %>% 
  getSignificant(design = ~MONTH, alpha = 0.01, number = 50, vst = TRUE) %>% # 
  plot_heatmap(taxa.order = "Phylum",
               taxa.label = "Genus",
               sample.label = "ID_friendly", 
               sample.order = "SORTED_genus")
plot
  
```


### 3.4 Get a DESeq2 transformed phyloseq object:

*getVst* For downstream analysis, it can be beneficial to use a transformed data-set. This function makes variance stabilizing transformation of the phyloseq object.

```{r eval=FALSE}
ps_16S %>% 
  subset_samples(SORTED_type=="Rotifer") %>% #Subsetting
  getVst(~SORTED_genus) %>% # DESeq2 transformation
  phyloseqToCSV(ID_friendly, "~/Desktop/rotifersvst_RSV.csv") # Continue with the result, or prin to CSV.
  
```


# 4. Other analysis.

### 4.1 Condense the phyloseq object to a matrix.

Many R-packages and plots require a simple matrix as an input object. (Visweb, Plotweb ect). The ´phyloseqToMatrix* function condenses the phyloseq object to a matrix with custom specified rownames and colnames.

```{r message=FALSE}
library(bipartite) # Install Bipartite if it is not installed.
```


```{r}

ps_18S %>% 
  subset_samples(SORTED_genus=="Keratella") %>%
  subset_taxa(Species != "Rotifera_XX_x_x") %>% 
  phyloseqToMatrix(colname="ID_friendly", rowname="Species") %>%
  plotweb(method="normal") # Plot from the Bipartite package.


```





# 5. Session info

```{r}
sessionInfo()
```
